Unity in diversity: Children from 17 diverse communities process gaze in similar ways

### 2024-10-13 paper/paper.Rmd
paper reformat step 1

```{r data}
data <- read_csv("../data/gafo-cc-clean-data.csv")%>%
  mutate(continent = ifelse(
    community %in% c("leipzig", "plymouth"), "Europe", 
    ifelse(
      community %in% c("akure", "hai||om", "khwe", "chimfunshi", "bandongo", "bayaka", "windhoek", "uganda"), "Africa",
      ifelse(community %in% c("stanford", "mexico", "buenos_aires"), "Americas",
             ifelse(community %in% c("india", "beijing", "turkey"), "Asia", "Oceania"))
    )
  ))%>%
  mutate(continent = factor(continent, levels = c("Americas", "Africa", "Europe", "Asia", "Oceania"), ordered = T))%>%
  mutate(community = recode(community,
                            akure = "Akure (Nigeria)",
                            leipzig = "Leipzig (Germany)",
                            `hai||om` = "Hai||om (Namibia)",
                            khwe = "Khwe (Namibia)",
                            windhoek = "Windhoek (Namibia)",
                            stanford = "Stanford (USA)",
                            chimfunshi = "Chimfunshi (Zambia)",
                            mexico = "Ocuilan (México)",
                            plymouth = "Plymouth (UK)",
                            beijing = "Beijing (China)", 
                            india = "Pune (India)",
                            buenos_aires = "Buenos Aires (Argentina)",
                            auckland = "Auckland (New Zealand)",
                            turkey = "Malatya (Türkiye)",
                            bandongo = "Bandongo (Rep. Congo)",
                            bayaka = "BaYaka (Rep. Congo)", 
                            uganda = "Nyabyeya (Uganda)"))
```

# Research Transparency Statement

All authors declare no conflicts of interest. Preregistration: The hypotheses, methods and parts of the analysis plan were preregistered (<!-- https://osf.io/tdsvc/?view_only=f404d66c967542f28ce4af16371f34e6-->https://osf.io/tdsvc) on March 12th, 2022, prior to data collection which began on March 18th, 2022. Additional analysis and deviations from the preregistration are reported in the Supplementary Material. Materials: All study materials are publicly available <!--[masked for peer review] --> (https://ccp-odc.eva.mpg.de/tango-cc/). Data: All primary data are publicly available <!--[masked for peer review; see .zip file included--> (https://github.com/ccp-eva/gafo-cc-analysis/). Analysis scripts: All analysis scripts are publicly available <!--[masked for peer review; see .zip file included]--> (https://github.com/ccp-eva/gafo-cc-analysis/).

@@

Human socio-cognitive skills enable unique forms of communication and cooperation that provide a bedrock for cumulative culture and the formation of complex societies  [@tomasello2020adaptive; @laland2021understanding; @wellman2014making; @henrich2016secret; @tomasello2003makes; @legare2019development; @heyes2018cognitive]. The eyes are the proverbial "window to the mind" and eye gaze is essential for many social reasoning processes  [@shepherd2010following; @doherty2006development; @emery2000eyes]. Others’ eye gaze is used to infer their focus of visual attention, which is a critical aspect of coordinated activities, including communication and cooperation [@tomasello2007reliance; @scaife1975capacity; @langton2000eyes; @rossano2012gaze; @richardson2005looking; @sebanz2006joint].

The ability to follow gaze emerges early in development [@tang2023slow; @gredeback2010development; @byers2021development; @del2019developmental]. The earliest signs of gaze following have been found in infants as young as four months [@astor2021gaze; @d1997demonstration]. Initially, infants rely more on head direction than actual gaze direction [@michel2021effects; @lempers1977development]. Throughout the first two years of life, children refine their abilities: they interpret gaze in mentalistic terms, for example, they follow gaze to locations outside their own visual field by moving around barriers [@moll200412]. Importantly, individual differences in children’s gaze following abilities predict later life outcomes, most notably communicative abilities [@carpenter1998social]. For example, gaze following at 10 months predicts language scores at 18 months of age [@brooks2005development]. Difficulties with gaze following have been linked to developmental disorders, including Autism [@itier2009neural; @thorup2016altered; @thorup2018reduced]. This work highlights the importance of gaze following as a foundational building block of human social interaction and its central place in theorizing.

A central assumption in the theoretical and empirical work discussed above is that, despite substantial variation in developmental contexts, gaze following works and develops in the same way across human societies [@tomasello2019becoming]. This assumption -- despite being central to many developmental theories -- is currently not supported by evidence. On the contrary, cross-cultural studies have revealed substantial diversity in socio-cognitive development [@miller2018contributions; @wellman2014making; @mayer2013synchrony; @dixson2018scaling; @taumoepeau2019cross]. One of the very few cross-cultural studies also found differences in the likelihood to follow gaze between communities [@callaghan2011early].

One potential source for this paradox lies in the reliance on aggregated measures in cross-cultural studies. Absolute differences in mean performance across communities are interpreted as a signal of different underlying cognitive processes. In the present study, we resolve this paradox by instead focusing on processing signatures that can be investigated independently of absolute community-level differences. This allows us to directly evaluate the empirical foundation of claims about universal features of human social cognition. To this end, we conducted a pre-registered, large-scale, cross-cultural study on the development of gaze following abilities to study potentially universal processing signatures.

The processing signatures were derived from a computational model that assumes that participants follow gaze by estimating a vector emanating from the eye center through the pupil [@prein2024variation]. The key innovation of the model is that it explains how individuals may use the same cognitive process but still differ in their measured abilities. The process always involves estimating a vector but also involves a degree of uncertainty because the eye center is not directly observable. Individuals are assumed to differ in their level of uncertainty with which they estimate the vector which causes differences in their observable behavior. Importantly, the assumed process leaves a key signature in the data that is observable independent of the absolute level of performance. In the present study, we therefore focus on this signature instead of absolute levels of performance when evaluating the claim whether there is evidence for a universal cognitive mechanism underlying gaze following.

The `r n_distinct(data$subjid)` participants who took part in the study lived in 17 different communities across 14 countries and five continents (Fig. \@ref(fig:fig1)A, Tab. \@ref(tab:tab1)). These countries represent ~46% of the world’s population. Communities covered a broad spectrum of geographical locations, social and political systems, languages, and subsistence styles (see Supplementary Material). This diversity allowed us to overcome the common pitfall of cross-cultural studies that compare urban communities from the Global North to rural communities from the Global South [@barrett2020towards].

@@

@@

A total of `r sum(as.numeric(demtab$N))` children between `r round(min(data$ageinyears), 2)` and `r round(max(data$ageinyears), 2)` years of age provided data for the study. Children lived in `r n_distinct(data$community)` different communities, located in `r n_distinct(demtab$country)` different countries. Table \@ref(tab:tab1) gives the sample size per community together with some basic demographic information. The recruitment strategy for each community is reported in the respective site descriptions. For some children, the exact birthday was unknown. In such cases, we set the birthday to the 30th of June of the year that would make them fall into the reported age category.

@@

## Setup and Procedure

The task was implemented as a browser-based interactive picture book using `HTML`, `CSS`, and `JavaScript`. Participants saw animated agents on a touch screen device, listened to pre-recorded audio instructions and responded by touching the screen. In all communities, a research assistant, fluent in the local language(s), guided the child through the task.

@@

The location (x-coordinate) where the balloon landed was determined in the following way: The screen was divided in ten equally sized bins. On each trial, one of the bins was randomly selected and the exact x-coordinate was randomly chosen within that bin. Constraints were that the balloon landed in each bin equally often and the same bin appeared no more than twice in a row.

@@

In addition to the gaze following task, caregivers responded to a short questionnaire about children’s access to screens and touchscreens (binary answer) as well as the number of people, children and children younger than the focal child living in the household (numeric; see Supplementary Material for details). The numeric variables were scaled within community prior to inclusion into the regression models.

@@

## Regression models

We used Bayesian Regression models fit in `R` [@r] using the package `brms` [@burkner2017brms] for all analyses except the cognitive models (see below). We used default priors built into `brms`. The dependent variable in all regression models was imprecision, that is, the absolute distance between the true location of the balloon (x-coordinate of its center) and the location where the participant touched the screen. We used a Log-normal distribution to model the data because the natural lower bound for imprecision is zero and the data was right skewed with a long tail. Numeric predictors that entered the models were scaled to have a mean of zero and a standard deviation of 1.

To analyse cross-cultural variation in performance, we used a cross-validation procedure [see e.g., @stengelin2023children]. In the Supplementary Material we give a detailed justification of this approach. For each community, we randomly sampled a data set that was 5/6 the size of the full data set (training data). Then, we fit the model to this training data and used the estimated model parameters to predict the remaining 1/6 of the data (testing data). We then compared the model predictions from the different models by computing the mean difference between the true and predicted imprecision, over all trials in the testing data set. We repeated the cross-validation procedure 100 times and computed the percentage of cases in which one model outperformed the other. We compared three models: a null model assuming no systematic community-level variation, a model assuming variation between communities and a model assuming variation between communities and in developmental trajectories (see Supplementary Material for model equations).

To evaluate the processing signatures predicted by the cognitive model that trials in which the balloon lands further away from the center lead to larger imprecision (see next section for details), we fit a model predicting imprecision by age and target centrality (distance of the landing position from the center in pixel/SVG units) with random intercepts for participant and community and random slopes for target centrality within participant and community (`brms` notation: `age + target_centrality + (target_centrality | participant) + (age + target_centrality | community)`).

## Cognitive model

Recent computational work modeled gaze following as social vector estimation [@prein2024variation]. When following gaze, onlookers observe the location of the pupil within the eye and estimate a vector emanating from the center of the eye through the pupil. The focus of attention is the location where the estimated vectors from both eyes hit a surface (Fig. \@ref(fig:fig12)). It is assumed that this estimation process has some uncertainty because the center of the eye is not directly observable and that individuals vary in their level of uncertainty. As a consequence, even though individuals use the same general process, they might differ in their absolute levels of precision. Crucially, this process model predicts a clear performance signature in the gaze following task: Trials in which the agent looks further away from the center should result in lower levels of precision compared to trials in which the agent looks closer to the center. This prediction is best understood by considering a similar phenomenon: pointing a torch light to a flat surface. The width of the light beam represents each individual’s level of uncertainty in vector estimation. When the torch is directed straight down, the light beam is concentrated in a relatively small area. When the torch is rotated to the side, the light from one half of the cone must travel further than the light from the other half to reach the surface. As a consequence, the light is spread over a wider area (see Fig. \@ref(fig:fig12)).

The model inversely models the process generating touches on the screen based on observed eye movements and is defined as:

@@

Here, $\theta$ represents an individual's cognitive ability to locate the focus of the agent's attention, $x_c$ represents the touched coordinate, and $\alpha_l$ and $\alpha_r$ correspond to the left and right pupil angles (each defined as the angle between a line connecting the center of the eye to the pupil and a line extended vertically downward from the center of the eye).

@@

$P(x_c)$ is a prior over potential target locations. Because the target was last visible in the screen and because the agent was located in the center, we assumed that participants have an a priori expectation that the target will land close to the middle. We estimated the strength of this center bias (*i.e.,* the standard deviation of a Normal distribution around the screen center) based on the data: $P(x_c) \sim \mathcal{N}(960, \sigma^p)$.

@@

Here, $\sigma_v$ determines the level of accuracy with which participants estimated the pupil angles, and it is thus the component of the model that defines $\theta$. Smaller values of $\sigma_v$ result in a narrow distribution around the pupil angle, making touches far away from the target less likely. Conversely, larger values for $\sigma_v$ lead to a wider distribution, making touches far away from the target more likely. To circle back to the analogy introduced above, $\sigma_v$ corresponds to the width of the light beam. Thus, the goal of the model was to estimate participant-specific values for $\sigma_v$: $\sigma_{v_i}$. For more details on how $\sigma_{v_i}$ was estimated, see the Supplementary Material.

To summarize, the model assumes that participant's touches are generated by a process that relies on noisy estimates of the agent's gaze direction. The precision, with which the gaze direction is estimated, varies between participants and increases with development.

As stated above, the key signature prediction of the model is that precision decreases when the balloon lands further away from the center. This pattern, however, also arises when participants ignore the agent's gaze completely and instead follow simple heuristics. We implemented these heuristics as alternative models and directly compared them to the gaze estimation model. According to the center bias model, participants always try to touch in the center of the screen: $P(x_c) \sim \mathcal{N}(960, 160)$ (960 is the x-coordinate of the center and 160 is the width of the balloon). According to the random guessing model, they randomly touch coordinates on the screen: $P(x_c) \sim \mathcal{U}(0, 1920)$.

The cognitive models were implemented in the probabilistic programming language `webppl` [@dippl]. All models were run separately for each community. Information on the prior distributions for all model parameters can be found in the associated online repository. We compared models based on the marginal likelihood of the data for each model, which represents the likelihood of the data while averaging over the prior distribution on parameters. The pair-wise ratio of marginal likelihoods for two models is known as the Bayes Factor. Bayes Factors are a quantitative measure of the predictive quality of a model, taking into account the possible values of the model parameters weighted by their prior probabilities. The incorporation of the prior distribution over parameters in the averaging process implicitly considers model complexity: models with more parameters typically exhibit broader prior distributions over parameter values and broader prior distribution can attenuate the potential gains in predictive accuracy that a model with more parameters might otherwise achieve [@lee2014bayesian].

@@

@@

>>>>

```{r data}
data <- read_csv("../data/gafo-cc-clean-data.csv")%>%
  mutate(continent = ifelse(
    community %in% c("leipzig", "plymouth"), "Europe", 
    ifelse(
      community %in% c("akure", "hai||om", "khwe", "chimfunshi", "bandongo", "bayaka", "windhoek", "uganda"), "Africa",
      ifelse(community %in% c("stanford", "mexico", "buenos_aires"), "Americas",
             ifelse(community %in% c("india", "beijing", "turkey"), "Asia", "Oceania"))
    )
  ))%>%
  mutate(continent = factor(continent, levels = c("Americas", "Africa", "Europe", "Asia", "Oceania"), ordered = T))%>%
  mutate(community = recode(community,
                            akure = "Akure (Nigeria)",
                            leipzig = "Leipzig (Germany)",
                            `hai||om` = "Hai||om (Namibia)",
                            khwe = "Khwe (Namibia)",
                            windhoek = "Windhoek (Namibia)",
                            stanford = "Stanford (USA)",
                            chimfunshi = "Chimfunshi (Zambia)",
                            mexico = "Ocuilan (México)",
                            plymouth = "Plymouth (UK)",
                            beijing = "Beijing (China)", 
                            india = "Pune (India)",
                            buenos_aires = "Buenos Aires (Argentina)",
                            auckland = "Auckland (New Zealand)",
                            turkey = "Malatya (Türkiye)",
                            bandongo = "Bandongo (Rep. Congo)",
                            bayaka = "BaYaka (Rep. Congo)", 
                            uganda = "Nyabyeya (Uganda)"))

data%>%
  mutate(side = ifelse(targetcenterx < 960, "left", "right"),
         error = ifelse(clickdistfromtargetcenterx < 0, "left", "right"))%>%
  group_by(side)%>%
  summarise(left_bias = mean(error == "left"),
            right_bias = mean(error == "right"))

@@

Human socio-cognitive skills enable unique forms of communication and cooperation that provide a bedrock for cumulative culture and the formation of complex societies [@tomasello2020adaptive; @laland2021understanding; @wellman2014making; @henrich2016secret; @tomasello2003makes; @legare2019development; @heyes2018cognitive]. The eyes are the proverbial "window to the mind" and eye gaze is essential for many social reasoning processes [@shepherd2010following; @doherty2006development; @emery2000eyes; @frischen2007gaze]. Others’ eye gaze is used to infer their focus of visual attention, which is a critical aspect of coordinated activities, including communication and cooperation [@tomasello2007reliance; @scaife1975capacity; @langton2000eyes; @rossano2012gaze; @richardson2005looking; @sebanz2006joint]. During ontogeny, gaze following is an important aspect of many critical learning objectives that enable children to become functioning members of the society they grow up in, including language, social learning and joint action [@brownell2011early; @moore2008development; @brooks2005development; @carpenter1998social; @mundy2007attention; @stephenson2021gaze]. Because of the central role gaze following plays during human ontogeny, it has been widely argued that gaze following has been a target of natural selection [@tomasello2007reliance; @kano2023evolution; @emery2000eyes; @clark2023white]. This implies that the process by which humans use gaze direction to infer the focus of attention is universal. In this paper, we report a comprehensive cross-cultural study on the ontogeny of gaze following in which we shed light on the universal aspects of gaze following as well as sources of variation and their origins.

## Ontogeny of gaze following

The ability to follow gaze emerges early in development [@tang2023slow; @gredeback2010development; @byers2021development; @del2019developmental]. The earliest signs of gaze following have been found in infants as young as four months [@astor2021gaze; @d1997demonstration]. Throughout the first two years of life, children refine their abilities: they interpret gaze in mentalistic terms [@butterworth1991minds; @deak2000effects], for example, they follow gaze to locations outside their own visual field by moving around barriers [@moll200412]. Initially, children rely more on head direction than actual gaze direction [@michel2021effects]. In fact, when head and gaze direction diverge, children fail to accurately locate the agent's focus of attention up until 19 months of age [@lempers1979young].

From an evolutionary perspective, while many species are able to follow gaze based on head directions, uniquely human forms of joint action and communication require a more precise localization of other's attention and thus critically rely on gaze direction inferred from eye movements [@emery2000eyes; @hessels2020does]. In a recent study, @prein2024variation studied the development of gaze following based on eye movements across from three years up until old age. They found particularly steep developmental improvements in the preschool years resulting in a relatively stable level of accuracy from ten years onward and a slight decrease starting around age 40.

The studies reported thus far, all relied on data collected in western affluent settings. Such settings represent only a minority of the worlds population and are thus insufficient to make claims about universal aspects of human cognition [@nielsen2017persistent; @norenzayan2005psychological; @amir2020cross]. Three studies with infants and children from traditionally underrepresented parts of the world (Bhutan, India, Peru, Vanuatu) find that children start gaze following (including head direction) at similar ages, the rates of gaze following, however, differed between cultural settings [@hernik2019infant; @astor2022maternal; @callaghan2011early].

Rates and accuracy of gaze following do not just differ between cultural settings; there is also substantial variation within settings. In fact, the pivotal role of gaze following in many uniquely human activities has been studied by relating individual differences in gaze following to other phenomena -- both cross-sectionally and longitudinally [@carpenter1998social; @brooks2015connecting]. For example, gaze following at 10 months predicts language scores at 18 months of age [@brooks2005development; see also @macdonald2013eye]. Furthermore, difficulties with gaze following have been linked to developmental disorders, including Autism [@itier2009neural; @thorup2016altered; @thorup2018reduced] and -- at least in some cultural contexts -- to maternal postpartum depression [astor2022maternal]. Individual differences are also key to explaining the -- mainly social-interactional -- driving forces behind the development of gaze following. For example, it has been found that early attachment quality or the use of gaze in communicative interactions predict later rates of gaze following [@astor2020social; @movellan2002development; @senju2015early].

## Cognitive universals and sources of variation

The existence of variation in gaze following both within and between cultural settings raises the question of how to square these findings with the suggestion that gaze following is a fundamental building block of human social cognition and interaction and that eye movements are processed the same way in humans all over the world. Looking at other aspects of social cognition, one could easily make the argument that variation is the norm rather than the exception [see e.g., @miller2018contributions; @wellman2014making; @mayer2013synchrony; @dixson2018scaling; @taumoepeau2019cross]. As a first step, answering this question requires data from many different cultural settings. In a second step, however, we need to a way of detecting universal processes in such data. The traditional approach is to compare some sort of aggregate measure (mean level of performance, average age of onset) across cultural settings. Absolute differences in mean performance across communities are interpreted as a signal of different underlying cognitive processes [see e.g., @house2020universal; @van2018development; @blake2015ontogeny; @kanngiesser2022children]. No differences seemingly support the existence of a psychological universal. Such an approach, however, neglects the existence of within-cultural variation altogether [see also @gurven2018broadening].

In the present study, we want to take a different approach for which we assume that universal processes and variation can co-exist [@kline2018variation; @greenfield2003cultural; @jensen2012bridging]. Instead of starting with the outcome (performance in the task), we start with the process that generates the outcome. By defining this process, we make a proposal for the universal aspect of the process. At the same time, we define variable aspects in the process that generate individual differences. This allows us to define signatures that the process leaves behind that can be detected independent of absolute levels of performance.

For gaze following, we can use the computational model proposed by @prein2024variation to derive such predictions. They formalized the widely-held view that gaze following involves estimating a vector emanating from the eye center through the pupil [@todorovic2006geometrical; @yaniv1990heuristics; @butterworth1991minds; @symons2004you]. The key innovation of the model is that it explains how individuals may use the same  process but still differ in their measured abilities. The process always involves estimating a vector but also involves a degree of uncertainty because the eye center is not directly observable. Individuals are assumed to differ in their level of uncertainty with which they estimate the vector which causes differences in their observable behavior. Importantly, the assumed process leaves a key signature in the data that is observable independent of the absolute level of performance. In the present study, we therefore focus on this signature instead of absolute levels of performance when evaluating the claim whether there is evidence for a universal cognitive process underlying gaze following.

## The current study

The present study had three goals. First, to collect a comprehensive data set and study the ontogeny of gaze following beyond infancy across cultures. To make this possible, we used a semi-standardized task that required minimal assistance from an experimenter and no behavioral coding. The task is an animated picture book presented on a tablet screen. Children watched a balloon disappear behind a hedge. An agent followed the trajectory of the balloon with their eyes (Figure \@ref(fig:fig1)B). The key dependent variable was (im)precision, that is, the deviation between where the agent looked (where the balloon was) and the child’s response. The task's flexible implementation as a browser-based web-app allowed us to quickly tailor its visual and audio content to each cultural setting (visuals and audio). The task has been psychometrically evaluated and has shown to yield reliable individual-level measurements across communities and across ages [@prein2024tango; @prein2024measuring].

We collected data in 17 different communities across 14 countries and five continents. Communities covered a broad spectrum of geographical locations, social and political systems, languages, and subsistence styles. This diversity allowed us to overcome a common pitfall of cross-cultural studies that compare urban communities from the Global North to rural communities from the Global South [@barrett2020towards]. We aimed for large sample sizes within each community to contrast within- and between cultural variation. Our expectation regarding the first goal was to see substantial variation across cultures but even more variation between individuals. In all communities, we expected performance to increase with age.

The second goals was to look for signatures in the data of the universal gaze following process predicted by the model of @prein2024measuring. In the task, the hidden object is located on a horizontal plane at the lower end of the screen. The agent is located in the upper center of the screen (see Figure \@ref(fig:fig1)B). The model predicts trials in which the object is hidden further away from the center to be more difficult resulting in higher imprecision. The signature is thus a u-shaped relation between object location and imprecision (Figure \@ref(fig:fig12)).

Finally, we sought to explain individual differences in gaze following precision by linking them to methodological aspects of the study as well as aggregate measures of children's everyday social experience. Experience with tablets and touch screens co-varied with community and we expected children more familiar with this medium to perform better. Previous work suggested that gaze following is refined in social interaction [@movellan2002development; @senju2015early]. To approximate social interaction, we asked parents to fill out a questionnaire about household size and composition. We acknowledge that this measure approximates opportunities for social interaction in a rather crude way but we nevertheless expected children living in larger households and with more siblings (relative to their community) to be more accurate when following gaze.

@@

## Preregistration

The study design, the sampling strategy and the general analytic strategy were preregistered prior to data collection (https://osf.io/tdsvc). The final sample size was not preregistered because we did not know how many communities would participate when the study began. Instead, we stated the age range we intended to study in each community (3.0 to 5.9 years of age) along and that we planned to test 20 children per year bin. We achieved this goal for most ages in most communities, but not for all (see Supplementary Table 1). The analysis reported here deviated from the preregistration in the following ways: In the pre-registration, in the regression models, we did not include random slopes for target centrality (i.e. the distance of the location where the balloon landed from the center of the screen), neither within subject nor within cultural setting. Instead, we included random slopes for trial within subject. We decided to change this and remove random slopes for trial but include them for target centrality because trial effects in the sense of learning across trials were unlikely because participants did not receive differential feedback. Furthermore, multiple studies with the same task since the registration found no trial effects [@prein2024tango; @prein2024variation]. On the other hand, we included random slopes for target centrality to be able to look at cross-cultural variation. The cognitive models were not mentioned in the pre-registration because it is a more recent development [@prein2024variation].

## Open data and materials

All study materials <!--[masked for peer review] --> (https://ccp-odc.eva.mpg.de/tango-cc/), primary data and analysis scripts are publicly available <!--[masked for peer review; see .zip file included--> (https://github.com/ccp-eva/gafo-cc-analysis/).

@@

A total of `r sum(as.numeric(demtab$N))` children between `r round(min(data$ageinyears), 2)` and `r round(max(data$ageinyears), 2)` years of age provided data for the study. Children lived in `r n_distinct(data$community)` different communities, located in `r n_distinct(demtab$country)` different countries across five continents. Table \@ref(tab:tab1) gives the sample size per community together with basic demographic information and age. For some children, the exact birthday was unknown. In such cases, we set the birthday to the 30th of June of the year that would make them fall into the reported age category. We provide a detailed description of the sample characteristics, the study site and recruitment strategy for each community in the Supplementary Material.

@@

## Material and Procedure

The task was implemented as a browser-based interactive picture book using `HTML`, `CSS`, and `JavaScript`. Participants saw animated agents on a touch screen device, listened to pre-recorded audio instructions and responded by touching the screen. In all communities, a research assistant, fluent in the local language(s), guided the child through the task. That is, the research assistant guided the child through the introduction and advanced the study from trial to trial.

@@

The location (x-coordinate) where the balloon landed was determined in the following way: The screen was divided in ten equally sized bins. On each trial, one of the bins was randomly selected and the exact x-coordinate was randomly chosen within that bin. Constraints were that the balloon landed in each bin once in the first ten trials and, for the remaining six test trials, it landed in a different om each trial. Thus, each bin appeared a no more than twice.

@@

In addition to the gaze following task, caregivers responded to a short questionnaire about children’s access to screens and touchscreens (binary answer) as well as the number of people, children and children younger than the focal child living in the household (numeric; see Supplementary Material for details).

@@

## Cross-cultural variation

We used Bayesian Regression models fit in `R` [@r] using the package `brms` [@burkner2017brms]. We used default priors built into `brms`. The dependent variable in all regression models was imprecision, that is, the absolute distance between the true location of the balloon (x-coordinate of its center) and the location where the participant touched the screen. We used a Log-normal distribution to model the data because the natural lower bound for imprecision is zero and the data was right skewed with a long tail. Numeric predictors that entered the models were scaled to have a mean of zero and a standard deviation of 1.

The first analysis was focused on cross-cultural variation. Fixed effects in the model were age and target centrality (distance of the landing position from the center in pixel/SVG units). The latter term accounts for trial difficulty. Furthermore, we included participant as a random effect, with a random slope for target centrality. To assess cross-cultural variation, we compared three models. A null model without cultural setting as a predictor (`brms` notation: `imprecision ~ age + target_centrality + (target_centrality | participant)`), a model with cultural setting as a random intercept (`imprecision ~ age + target_centrality + (target_centrality | participant) + (target_centrality | community)`) and a model with cultural setting as a random intercept and an added random slope for age (`imprecision ~ age + target_centrality + (target_centrality | participant) + (age + target_centrality | community)`). Thus, the second model assumes that there is variation across cultures in average levels of precision and the third model assumes that there are additional cultural differences in the effect of age.

As stated in the pre-registration, comparing these models could be problematic. Participants are fully nested within cultural setting. If there was an effect of cultural setting, we would expect participant random intercepts to cluster by cultural setting. This clustering would appear whether or not cultural setting would be included in the model as a random effect or not -- the only difference would be if the participant random intercepts were estimated as a deviation from a grand intercept or a culture-specific one. Standard metrics such as `WAIC` or `LOO` would penalize the model with additional intercept for cultural setting for having additional parameters that do not help to improve predictive accuracy.

To get around this problem, we used a cross-validation procedure (see e.g., 6). For each cultural setting, we randomly sampled a data set that was 5/6 the size of the full data set (training data). Then, we fit the model to this training data and used the estimated model parameters to predict the remaining 1/6 of the data (testing data). We then compared the model predictions from the different models by computing the mean difference between the true and predicted imprecision, over all trials in the testing data set. This approach gets around the problem mentioned above because the model predicts a new data set for which the individual random intercepts are unknown. Clustering by culture could therefore only be predicted by a model that included culture as a predictor. We repeated the cross-validation procedure 100 times and counted which model performed best most often.

## Processing signatures

The processing signatures were derived from the model proposed by [@prein2024variation]. The model sees gaze following as social vector estimation. When following gaze, onlookers observe the location (and movement) of the pupil within the eye and estimate a vector emanating from the center of the eye through the pupil. The focus of attention is the location where the estimated vectors from both eyes hit a surface (Fig. \@ref(fig:fig12)). It is assumed that this estimation process is not perfect but has some uncertainty because the center of the eye is not directly observable. Individual differences are conceptualized as differences in the level of uncertainty. As a consequence, even though individuals use the same general process, they might differ in their absolute levels of precision.

The process model predicts a clear performance signature in the data: trials in which the agent looks further away from the center should result in lower levels of precision compared to trials in which the agent looks closer to the center. This prediction is best understood by considering a similar phenomenon: pointing a torch light to a flat surface. The width of the light beam represents each individual’s level of uncertainty in vector estimation. When the torch is directed straight down, the light beam is concentrated in a relatively small area. When the torch is rotated to the side, the light from one half of the cone must travel further than the light from the other half to reach the surface. As a consequence, the light is spread over a wider area (see Fig. \@ref(fig:fig12)).

In the following, we give a brief mathematical description of the model. The model inversely models the process generating touches on the screen based on observed eye movements and is defined as:

@@

Here, $\theta$ represents an individual's level of precision in locating the focus of the agent's attention, $x_c$ represents the touched coordinate, and $\alpha_l$ and $\alpha_r$ correspond to the left and right pupil angles (each defined as the angle between a line connecting the center of the eye to the pupil and a line extended vertically downward from the center of the eye).

@@

$P(x_c)$ is a prior over potential target locations. Because the target was last visible in the screen and because the agent was located in the center, we assumed that participants have an a priori expectation that the target will land closer to the middle. We estimated the strength of this center bias (i.e., the standard deviation of a Normal distribution around the screen center) based on the data: $P(x_c) \sim \mathcal{N}(960, \sigma^p)$.

@@

Thus, $\sigma_v$ determines the level of accuracy with which participants estimated the pupil angles, and it is thus the component of the model that defines $\theta$. Smaller values of $\sigma_v$ result in a narrow distribution around the pupil angle, making touches far away from the target less likely. Conversely, larger values for $\sigma_v$ lead to a wider distribution, making touches far away from the target more likely. To circle back to the analogy introduced above, $\sigma_v$ corresponds to the width of the light beam. Thus, the goal of the model was to estimate participant-specific values for $\sigma_v$: $\sigma_{v_i}$. For more details on how $\sigma_{v_i}$ was estimated, see the Supplementary Material.

As stated above, the key signature prediction of the model is that precision decreases when the balloon lands further away from the center. To test this prediction, we fit a model predicting imprecision by age and target centrality with random intercepts for participant and community and random slopes for target centrality within participant and community (`imprecision ~ age + target_centrality + (target_centrality | participant) + (age + target_centrality | community)`). As stated above, the predictor target centrality captures the distance from the center so that a positive effect of target centrality (i.e. a positive estimate with a 95% CrI not overlapping with zero) would mean support for the processing signature. In addition, we visualized the data for each community and inspected the shape of the plot.

The same pattern, however, also arises when participants ignore the agent's gaze completely and instead follow simple heuristics. When participants always touch the center of the screen, regardless of where the agent is looking, trials in which the balloon lands further away from the center have a higher imprecision. When participants randomly touch a location on the screen -- again ignoring the agent's gaze -- the maximum imprecision for trials in which the balloon lands in the center is half the width of the screen. When the balloon lands on one of the far ends of the screen, the maximum imprecision is a full screen widths. Thus, across trials, the average imprecision is again higher when the balloon lands further away from the center, resulting in the same pattern as predicted by the model.

Even though these alternatives are unlikely because they assume that participants ignore the agent's gaze, we nevertheless want to rule them out as processes generating the data. Thus, we implemented the gaze model along with the two alternative models in the probabilistic programming language `webppl` [@dippl]. The way the gaze model predicts the participants behavior has been described above. The center bias model predicts a participant's touch by sampling from a Normal distribution around the center of the screen $P(x_c) \sim \mathcal{N}(960, 160)$ (960 is the x-coordinate of the center and 160 is the width of the balloon). The random guessing model predicts participant's touch by sampling from a uniform distribution over all possible locations: $P(x_c) \sim \mathcal{U}(0, 1920)$. Information on the prior distributions for all model parameters can be found in the associated online repository.

For each community, we compared models based on the marginal likelihood of the data for each model, which represents the likelihood of the data while averaging over the prior distribution on parameters. The pair-wise ratio of marginal likelihoods for two models is known as the Bayes Factor. Bayes Factors are a quantitative measure of the predictive quality of a model, taking into account the possible values of the model parameters weighted by their prior probabilities. The incorporation of the prior distribution over parameters in the averaging process implicitly considers model complexity: models with more parameters typically exhibit broader prior distributions over parameter values and broader prior distribution can attenuate the potential gains in predictive accuracy that a model with more parameters might otherwise achieve [@lee2014bayesian].

The final analysis focused on whether we could predict performance in the task by methodological aspects of the study and aggregate measures of everyday social experience. For the ease of model fitting, we aggregated the data for each participant so that models predicted the average imprecision across trials. This approach is justified because the mean is nearly perfectly correlated with a model-based estimate of a participant's ability [$\sigma_v$ in the model above, see @prein2024variation] and because the predictor variables did not vary within child.

In the questionnaire, we asked about children’s exposure to screens as well as touchscreens. These two variables were largely redundant and so we included only one of them in the model. We chose the availability of a touchscreen as a predictor because the task itself was presented on a touchscreen.

For household composition, we asked for the total number of people in the household, the number of children and the number of younger children. We standardized each predictor within each community before fitting the models. Thus, the interpretation of the coefficient is the gain in precision for living e.g., in a larger household relative to other children from the same community.

We compared a null model (`mean_imprecision ~ age + (age | community)`) to a model including access to touchscreens only as a fixed effect (`mean_imprecision ~ touchscreen + age + (age | culture)`), a model in which the effect of access to touchscreens was also allowed to vary by community (`mean_imprecision ~ touchscreen + age + (touchscreen + age | culture)`) and a model for each of the household-based predictors (e.g., `mean_imprecision ~ household_size + touchscreen + age + (age | culture)`). Models were fit in `brms` and compared based on the difference in expected log pointwise predictive density (`ELPD`) computed via the widely applicable information criterion (`WAIC`) and the standard error of that difference (SE). We inspected the estimates for fixed effects in the winning model along with their 95% CrI.

For a community-level perspective, we correlated the proportion of children with access to touchscreens with an age-corrected performance average for each community. To obtain the latter, extracted the random intercept estimates for community from the null model described above. Because the model also includes age as a fixed effect, these values reflect variation between communities once differences in age have been accounted for.

@@

```{r}
pred_model_comp <- readRDS("../saves/model_comp.rds")%>%
  mutate_if(is.numeric, format, digits = 2, nsmall = 2)%>%
  select(model, elpd_diff, se_diff, waic, se_waic, value)%>%
  mutate(model = recode(model, 
                        bm_pred_null = "null",
                        bm_pred_tc = "touchscreen",
                        bm_pred_tc_hh = "household",
                        bm_pred_tc_c = "children",
                        bm_pred_tc_yc = "younger children",
                        bm_pred_tc_rand = "touchscreen (by culture)"))
```

@@

Cultural settings are not homogeneous clusters but greatly overlap.

alternative models not really alternative processes - motor noise model predicts no u-shaped pattern.

The effect of touchscreen exposure wuggests that familiarity with the device used for data collection explains variation between communities. Children with more touchscreen experience were probably better at task handling and thus more likely to precisely touch the location they inferred the agent to look at. However, there was substantial variation between individuals that could not be explained by differential exposures to touchscreens alone. For example, in Malatya (Türkiye) where 100% of children had access to touchscreens there was still substantial variation between individuals (see Fig.\@ref(fig:fig1)B). This strongly indicates that other factors likely contributed to individual differences.
